---
---
title: "Linear Model LM - part I"
author: "GWU Intro to Data Science DATS 6101"
# date: "today"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---
---

#Reading CSV file and Displaying Data Dimensions

```{r} 
#Loading libraries
library(tidyverse)
library(ggthemes)
library(ggplot2) #for data visualization
library(dplyr) #for data manipulation
library(tidyr) #for tidying the data
library(corrgram)
library(lubridate)
library(caret)
library(corrplot)

#Reading the CSV file into data frame
train_data <- read.csv('train.csv', stringsAsFactors = FALSE)
test_data <- read.csv("test.csv", stringsAsFactors =  FALSE)
 
## Getting Data Frame dimensions
#displaying shape, number of columns, rows and first few rows of the data set
head(train_data)
nrow(train_data)  # Number of rows, 10866
ncol(train_data)  # Number of columns, 12

#Basic information about Data Frame
str(train_data)
```

```{r summary}
summary(train_data) #summary statistics for the data frame
```


# ATTRIBUTES OF DATA:
- date_time: records time and date
- season : season (1:spring, 2:summer, 3:fall, 4:winter)
- holiday : whether day is holiday or not 
- weekday : day of the week
- workingday : if day is neither weekend nor holiday is 1, otherwise is 0.
+ weathersit : 
    - 1: Clear, Few clouds, Partly cloudy
    - 2: Mist + Cloudy, Mist + Broken clouds, Mist + Few clouds, Mist
    - 3: Light Snow, Light Rain + Thunderstorm + Scattered clouds, Light Rain + Scattered clouds
    - 4: Heavy Rain + Ice Pallets + Thunderstorm + Mist, Snow + Fog
- temp : Normalized temperature in Celsius. The values are derived via (t-t_min)/(t_max-t_min), t_min=-8, t_max=+39 (only in hourly scale)
- atemp: Normalized feeling temperature in Celsius. The values are derived via (t-t_min)/(t_max-t_min), t_min=-16, t_max=+50 (only in hourly scale)
- hum: Normalized humidity. The values are divided to 100 (max)
- windspeed: Normalized wind speed. The values are divided to 67 (max)
- casual: count of casual users
- registered: count of registered users
- cnt: count of total rental bikes including both casual and registered

### The goal of normalization is to transform features to be on a similar scale. This improves the performance and training stability of the model.

#  Conducting Univariate Analysis 

### Plotting Continuous Data, Temperature

```{r plots, echo='FALSE'}
ggplot(train_data,aes(x=temp))+
  geom_histogram(binwidth = 0.02, fill = "blue",color="black")+
  geom_density(aes(y = after_stat(scaled)), color = "red")+
  labs(title = "Distribution of Temperature") +
  xlab("Temperature") +
  ylab("density")
```

```{r plot1, echo='FALSE'}
ggplot(train_data, aes(temp, count)) + geom_point(alpha = 0.3, aes(color = temp)) + theme_bw()
```

Observations:
- Every time a bike is shared, the temperature is noted
- Most people won't rent a bike when it is too hot or too cold. More bikes are rented when the temperature is pleasant.
- A high number of bikes are rented when the temperature is around 16 degree celsius and 26 degree celsius.

```{r, include='FALSE'}
ggplot(train_data, aes(registered, count))+ geom_point(aes(color= humidity)) + theme_light() +
  labs(x = "Register bike Users", 
       y = "Number of Bike Users",
       title = "Register Bike Users Vs Number of bike users")

```

#### Box_Plot gives us the following information:
- Minimum
- First Quartile
- Median
- Third Quartile
- Maximum
```{r boxplot, echo='FALSE'}
library(ggplot2)

#creating box plot for "temp" column
ggplot(train_data,aes(y=temp))+
  geom_boxplot(fill="blue",color="black")+
  labs(title = "Boxplot for temperature")+
  ylab("Temperature")
```
- Observation : Outliers are 1.5*IQR more than 3rd Quartile or 1.5*IQR less than 1st Quartile 

#### Plotting Categorical Variables 

### Instances of bike sharing across seasons

- Getting unique values in the 'season' column

```{r categorical_variable, echo= 'FALSE'}
library(ggplot2)
train_data$season <- as.factor(train_data$season)
unique(train_data$season) #gives us an array of [1,2,3,4]
#finding the count of the array of seasons
# 1: Spring , 2:Summer, 3: Fall, 4:Winter
table(train_data$season)
#based on the output here are the counts:
# 1: 2686 , 2: 2733 , 3: 2733, 4: 2734
ggplot(train_data, aes(x = factor(season), fill = factor(season))) +
  geom_bar() +
  labs(
    title = "Bike Sharing Across Seasons",
    x = "Season",
    y = "Count",
    fill = "Season"
  ) +
  scale_fill_brewer(palette = "Set2")
```
## Conducting Bivariate Analysis

### Plotting Continuous Data vs Continuous Data

### Are Temp and ATemp related ? Is there relationship between the two ?

```{r Continuous Data vs Continuous Data, echo= 'FALSE'}
library(ggplot2)

#Create a joint scatter plot for 'temp' vs 'atemp'
plot <- ggplot(train_data, aes(x=temp, y=atemp))+
  geom_point()+
  labs(title="Temp vs Atemp")

print(plot)

# Working on Correlation
correlation <- cor(train_data$temp, train_data$atemp)
#rounding the correlation coefficient to two decimal places
rounded_correlation <- round(correlation,2)
#printing the rounded_correlation 
cat("Correlation between 'temp' and 'atemp': ", rounded_correlation)
```
Observation: 
- Temp and atemp have a positive linear relation
- Correlation is 0.99
- 0 correlation doesn't mean no relation, it means no linear relation
- We would need to see the plot as well and not rely simply on correlation

### Plotting Categorical vs Continuous Variable

##### Boxplots of Continuous Variable over the Categorical Variable

### Plotting number of bikes rentend across four seasons

```{r Categorical Data vs Continuous Data, echo= 'FALSE'}
library(ggplot2)

#Create a joint scatter plot for 'temp' vs 'atemp'
train_data$season <- as.factor(train_data$season)

# Create a box plot for 'cnt' by 'season'
ggplot(train_data, aes(x = season, y = count, fill = season, group = season)) +
  geom_boxplot() +
  labs(title = "Total # Bikes Rented By Seasons") +
  scale_fill_brewer(palette = "Set2") +
  xlab("Season") +
  ylab("Total # Bikes Rented")

```
Observation:
- The cnt column is an important column for analysis and the outliers are good for the business because we are working with Bike Sharing Data. So, we will not treat them.





